---
title: "Multiple Linear Regression"
author: "Allison Xia"
date: "2023-12-06"
output: github_document
fontsize: 12pt
header-includes: \usepackage{setspace}\doublespacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE,warning=FALSE)
library(tidyverse)
library(dplyr)
library(patchwork)
library(corrplot)
```

```{r load data}
df = read_csv("data/Project_1_data.csv")|>
  janitor::clean_names()|>
  mutate(
    wkly_study_hours=ifelse(wkly_study_hours=="10-May","5-10",wkly_study_hours)
  )|>
  na.omit()
```



```{r categorical variables to factors}
# Transforming categorical variables to factors
df_transformed <- df |> 
  mutate(
    gender = as.factor(gender),
    ethnic_group = as.factor(ethnic_group),
    parent_educ = factor(parent_educ,
                         levels= c("some high school", "high school", "associate's degree", "some college", "bachelor's degree", "master's degree")),
    lunch_type = as.factor(lunch_type),
    test_prep = as.factor(test_prep),
    parent_marital_status = as.factor(parent_marital_status),
    practice_sport = factor(practice_sport,
                            levels = c("never", "sometimes", "regularly")),
    is_first_child = factor(is_first_child),
    transport_means = as.factor(transport_means),
    wkly_study_hours = factor(wkly_study_hours,
                              levels = c("< 5", "5-10", "> 10"))
  )

```

```{r MLR models}
# Build the MLR model for Math scores
model_math <- lm(math_score ~ gender + ethnic_group + parent_educ + lunch_type + test_prep + parent_marital_status + practice_sport + is_first_child + nr_siblings + transport_means + wkly_study_hours, data = df_transformed)
model_read <- lm(reading_score ~ gender + ethnic_group + parent_educ + lunch_type + test_prep + parent_marital_status + practice_sport + is_first_child + nr_siblings + transport_means + wkly_study_hours, data = df_transformed)
model_write <- lm(writing_score ~ gender + ethnic_group + parent_educ + lunch_type + test_prep + parent_marital_status + practice_sport + is_first_child + nr_siblings + transport_means + wkly_study_hours, data = df_transformed)
```

## MLR - Math
```{r math}
summary(model_math)
```

### Coefficients and Significance Levels:
* Intercept (44.1006): The expected value of math_score when all other predictors are at their reference level or zero.
* gendermale (5.0855, p < 0.001): Being male is associated with an average increase of 5.0855 points in math_score compared to females, holding all else constant. This is statistically significant.
* ethnic_group: Only ethnic_groupgroup E (11.1752, p < 0.001) is significant, suggesting students in this group score higher in math compared to the reference group.
* parent_educ: The associate's degree (4.9058, p = 0.00584), bachelor's degree (6.6652, p = 0.00140), and master's degree (6.8096, p = 0.00760) are significant and associated with higher math scores compared to the reference category.
* lunch_typestandard (12.3539, p < 0.001): Students with standard lunch type score significantly higher.
* test_prepnone (-4.7717, p < 0.001): Not participating in test preparation is associated with lower math scores.
* parent_marital_status: Married (5.4805, p = 0.00075) and Widowed (7.7944, p = 0.04134) are associated with higher scores.
* practice_sport: Not significant.
* is_first_childyes: Not significant.
* nr_siblings (0.7403, p = 0.05461): A borderline significant positive association with math scores.
* transport_meansschool_bus: Not significant.
* wkly_study_hours: Studying 5-10 hours (3.5394, p = 0.00863) shows a significant positive effect.

### Residuals:
The spread of residuals suggests the errors are somewhat symmetrically distributed around the predicted values, which is a good sign for linear regression assumptions.

### Model Fit:
Residual Standard Error (13.52): Indicates the average difference between the observed values and the values predicted by the model.

### R-squared:
* Multiple R-squared (0.3221): About 32.21% of the variability in math_score is explained by the model.
* Adjusted R-squared (0.2956): Adjusts the R-squared for the number of predictors, a better measure for models with multiple predictors.

### Statistic & p-value
F-statistic (12.18) and p-value (< 2.2e-16): The model is statistically significant, meaning it performs better than a model with no predictors.

```{r}
summary(model_read)
```

```{r}
summary(model_write)
```



